Geometric models overestimate lake depth due to imperfect slope measurement

### 2020-11-02 manuscript/agujournaltemplate.tex
initial minor text revisions

% The Plain Language Summary should be written for a broad audience,
% including journalists and the science-interested public, that will not have
% a background in your field.
%
% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
% see http://sharingscience.agu.org/creating-plain-language-summary/)
%
%% ------------------------------------------------------------------------ %%
%TC:endignore
%% \begin{abstract} starts the second page
\begin{abstract}
Lake depth is a critical characteristic that influences many important ecological processes in lakes. Unfortunately, lake depth measurements are labor-intensive to gather and are only available for a small fraction of lakes globally. Therefore, scientists have tried to predict lake depth from characteristics that are easily obtained for all lakes such as lake surface area or the slope of the land surrounding a lake. One approach for predicting lake depth simulates lake basins using a geometric model where nearshore land slope is assumed to be a representative proxy for in-lake slope and the distance to the center of the lake is assumed to be a representative proxy for the distance to the deepest point of the lake. However, these assumptions have rarely been tested in a broad range of lakes. We used bathymetry data from approximately 5,000 lakes and reservoirs to test these assumptions and to examine whether differences in lake type or shape influences depth prediction error. We found that nearshore land slope was not representative of in-lake slope and using it for prediction increases error substantially relative to models using true in-lake slope. We also found that models using nearshore land slope as a proxy systematically overpredict lake depth in concave lakes (i.e. bowl-shaped; up to 18\% of lakes in the study population) and reservoir lakes (up to 30\% of lakes in the study population), suggesting caution in using geometric models for depth prediction in unsampled lakes.
\end{abstract}

@@

\section{Introduction}
\noindent
\noindent
Lake depth is an important factor controlling lake physics, chemistry, and biota. Deeper lakes generally have higher water clarity and less complete mixing compared to shallow lakes \cite{feeEffectsLakeSize1996, readSimulating2368Temperate2014}. These differences are reflected in variation among lakes in terms of biological productivity \cite{qinWaterDepthUnderpins2020} and rates of greenhouse gas production \cite{liSignificantContributionLake2020}. However, because measured depth data is only available for a small fraction ($\sim$25\%) of all lakes, our ability to understand and predict depth-dependent processes is limited. The importance of lake depth, coupled with its limited availability, has led to numerous attempts to predict depth using measures available for all lakes such as lake surface area or the nearshore slope of the land surrounding a lake \cite{heathcotePredictingBathymetricFeatures2015, oliver2016prediction, sobekPredictingDepthVolume2011}. Given the limited prediction accuracy of prior depth prediction efforts ($\pm$ 6-7 m), a major focus has been on  improving prediction accuracy using strategies such as employing more diverse covariates \cite{oliver2016prediction}, varying lake buffer sizes \cite{heathcotePredictingBathymetricFeatures2015}, or estimating hidden groupings (e.g. fitting different models for distinct size classes) among lakes \cite{sobekPredictingDepthVolume2011}.

>>>>

% The Plain Language Summary should be written for a broad audience,
% including journalists and the science-interested public, that will not have
% a background in your field.
%
% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
% see http://sharingscience.agu.org/creating-plain-language-summary/)
%
%% ------------------------------------------------------------------------ %%
%TC:endignore
%% \begin{abstract} starts the second page
\begin{abstract}
Lake depth is a critical characteristic that influences many important ecological processes in lakes. Unfortunately, lake depth measurements are labor-intensive to gather and are only available for a tiny fraction of lakes globally. Therefore, scientists have tried to predict lake depth from characteristics that are easily obtained for all lakes such as lake surface area or the slope of the land surrounding a lake. One approach for predicting lake depth simulates lake basins using a geometric model where nearshore land slope is assumed to be a representative proxy for in-lake slope and the distance to the center of the lake is assumed to be a representative proxy for the distance to the deepest point of the lake. However, these assumptions have rarely been tested in a broad range of lakes. We used bathymetry data from approximately 5,000 lakes and reservoirs to test these assumptions and to examine whether differences in lake type or shape influences depth prediction error. We found that nearshore land slope was not representative of in-lake slope and using it for prediction increases error substantially relative to models using true in-lake slope. We also found that models using nearshore land slope as a proxy systematically overpredict lake depth in concave lakes (i.e. bowl-shaped; up to 18\% of lakes in the study population) and reservoir lakes (up to 30\% of lakes in the study population), suggesting caution in using geometric models for depth prediction in unsampled lakes.
\end{abstract}

@@

\section{Introduction}
\noindent
\noindent
Lake depth is an important factor controlling lake physics, chemistry, and biota. Deeper lakes generally have higher water clarity and less complete mixing compared to shallow lakes \cite{feeEffectsLakeSize1996, readSimulating2368Temperate2014}. These differences are reflected in variation among lakes in terms of biological productivity \cite{qinWaterDepthUnderpins2020} and rates of greenhouse gas production \cite{liSignificantContributionLake2020}. However, because measured depth data is only available for a tiny fraction of lakes ($\sim$25\% of all lakes in our study footprint), our ability to understand and predict depth-dependent processes is limited. The importance of lake depth, coupled with its limited availability, has led to numerous attempts to predict depth using measures available for all lakes such as lake surface area or the nearshore slope of the land surrounding a lake \cite{heathcotePredictingBathymetricFeatures2015, oliver2016prediction, sobekPredictingDepthVolume2011}. Given the limited prediction accuracy of prior depth prediction efforts ($\pm$ 6-7 m), a major focus has been on  improving prediction accuracy using strategies such as employing more diverse covariates \cite{oliver2016prediction}, varying lake buffer sizes \cite{heathcotePredictingBathymetricFeatures2015}, or estimating hidden groupings (e.g. fitting different models for distinct size classes) among lakes \cite{sobekPredictingDepthVolume2011}.
